\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-11-08 k-planar.tex
Just noticed issue with definitions

For any integer $d\ge 3$, an embedded (multi)graph $G$ is \defin{$d$-framed} if it has a biconnected plane spanning (multi)subgraph $G_0$, whose facial cycles each have length at most $d$,  and such that the interior of each edge in $E(G)\setminus E(G_0)$ is contained in the interior of some face of $G_0$. The embedded (multi)graph $G_0$ is called the \defin{frame} of $G$. This definition (for simple graphs) was introduced by \citet{BDGGMR}.

Every $d$-framed multigraph can be described by a $(2,d(d-3)/2)$-shortcut system applied to a plane multigraph by adding a vertex inside each face $F$ of $G_0$ adjacent to the vertices of $F$ and creating a shortcut between each non-adjacent pair of vertices in $V(F)$. Thus \cref{ShortcutProduct} is applicable. However, for framed graphs we prove the following stronger result, where the treewidth bound on $H$ is best possible.\footnote{A multigraph $G$ is \defin{contained} in a graph $X$ if the simple graph underlying $G$ is contained in $X$.}

@@

For readers already familiar with the proof in \cite{dujmovic.joret.ea:planar}, the main difference here is that a typical tripod in the proof of \cite[Lemma~17]{dujmovic.joret.ea:planar} consists of three upward paths in a BFS tree/forest whose lower endpoints belong to a single triangular face $\tau$.  In the current setting, a typical tripod consists of three upward paths belonging to a common face $\tau$ of $G_0$ plus all of the (at most $d$) vertices of $\tau$.

@@

\note{PM}{Here we're using $f$ to denote a face, earlier we were using $F$.}
If $G_0$ has a face $f$ of length 2, then add a new vertex inside $f$ adjacent to both vertices on the boundary of $f$, which creates two new triangular faces $f_1$ and $f_2$. If $f$ is a lake, then make $f_1$ and $f_2$ lakes. If $F$ is a nation, then make $f_1$ a nation and make $f_2$ a lake. The resulting map graph is still $G$. So we may assume that $G_0$ is an edge-maximal multigraph with no faces of length 2 (and with each face labelled a nation or a lake), such that $G$ is the corresponding map graph. This is well-defined since the assumption of having no face of length 2 implies that $|E(G_0|\leq 3|V(G_0)|-6$.

Suppose that some face $f$ of $G_0$ has a disconnected boundary (implying that $G_0$ is disconnected). Let $v$ and $w$ be vertices in distinct components of the boundary of $f$. Add the edge $vw$ to $G_0$ across $f$. The corresponding map graph is unchanged, which contradicts the edge-maximality of $G_0$. Thus no face of $G_0$ is disconnected.

Suppose that some face $f$ of $G_0$ has a repeated vertex $v$ in the boundary walk of $f$ and let $u,v,w$ be consecutive vertices in this boundary walk.  Since $v$ appears more than once in the boundary walk of $f$, the degree of $v$ is greater than $1$.  Since the degree of $v$ is greater than $1$ and the length of $f$ is greater than $2$, $u\neq w$. Since $v$ is repeated, $uw$ is not an edge in the boundary of $f$. Add the edge $uw$ across $f$ and consider the resulting face $uvw$ to be a lake. Since $v$ appears elsewhere in the boundary of $f$, the corresponding map graph is unchanged, which contradicts the edge-maximality of $G_0$. Thus no facial walk of $G_0$ has a repeated vertex. Since each facial walk is connected, each face of $G_0$ is bounded by a cycle, and $G_0$ is biconnected.

@@

Let $G_1$ be a plane supergraph of $G_0^*$ obtained by adding the nation cycle of each face $F$ of $G_0$, and then triangulating each face with more than $d$ vertices on its boundary. $G_1$ is biconnected since $G^*_0$ is a biconnected spanning subgraph of $G_1$. Each nation cycle of length at least $3$ is now a facial cycle of $G_1$, and no face of $G_1$ has more than $d$ vertices on its boundary. Let $\widehat{G}_1$ be the \note{DW}{Add ``edge-maximal''. Do we need to write this anywhere else?} $d$-framed graph whose frame is $G_1$.

>>>>

For any integer $d\ge 3$, an embedded (multi)graph $G$ is \defin{$d$-framed} if it has a biconnected plane spanning (multi)subgraph $G_0$, whose facial cycles each have length at most $d$,  and such that the interior of each edge in $E(G)\setminus E(G_0)$ is contained in the interior of some face of $G_0$. The embedded (multi)graph $G_0$ is called the \defin{frame} of $G$. This definition (for simple graphs) was introduced by \citet{BDGGMR}.  A $d$-framed graph $G$ is \defin{edge-maximal} with respect to its frame $G_0$ if, for each face $F$ of $G_0$ and each pair of distinct vertices $v,w\in V(F)$, $vw\in E(F)$ or $G$ contains an edge $vw$ through $F$.

Every edge-maximal $d$-framed multigraph can be described by a $(2,d(d-3)/2)$-shortcut system applied to a plane multigraph by adding a vertex inside each face $F$ of $G_0$ adjacent to the vertices of $F$ and creating a length-$2$ shortcut between each non-adjacent pair of vertices in $V(F)$. Thus \cref{ShortcutProduct,PlanarProduct} imply that any $d$-framed multigraph $G$ is a subgraph of $H\boxtimes P\boxtimes K_{O(d^2)}$ where the treewidth of $H$ is at most $10$.  However, for framed graphs we prove the following stronger result, where the treewidth bound on $H$ is best possible and dependence on $d$ is only linear.\footnote{A multigraph $G$ is \defin{contained} in a graph $X$ if the simple graph underlying $G$ is contained in $X$.}

@@

For readers already familiar with the proof in \cite{dujmovic.joret.ea:planar}, the main difference here is that a typical tripod in the proof of \cite[Lemma~17]{dujmovic.joret.ea:planar} consists of three upward paths in a BFS tree/forest whose lower endpoints belong to a single triangular face $\tau$.  In the current setting, a typical tripod consists of three upward paths whose lower endpoints belong to a common face $\tau$ of $G_0$ plus all of the (at most $d$) vertices of $\tau$ that are not already included in tripods.

@@

If $G_0$ has a face $F$ of length 2, then add a new vertex inside $F$ adjacent to both vertices on the boundary of $F$, which creates two new triangular faces $F_1$ and $F_2$. If $F$ is a lake, then make $F_1$ and $F_2$ lakes. If $F$ is a nation, then make $F_1$ a nation and make $F_2$ a lake. The resulting map graph is still $G$. We may therefore assume that $G_0$ has no face of length $2$.  We may also assume that $G_0$ is edge-maximal in the following sense: Adding any edge to $G_0$ will introduce a crossing or make it impossible to label the faces of $G_0$ as nations or lakes to obtain $G$ as the resulting map graph. This edge-maximal graph is well-defined since the assumption of having no face of length $2$ implies that $|E(G_0|\leq 3|V(G_0)|-6$.

We now argue that $G_0$ is biconnected.  Suppose that some face $F$ of $G_0$ has a disconnected boundary (implying that $G_0$ is disconnected). Let $v$ and $w$ be vertices in distinct components of the boundary of $F$. Add the edge $vw$ to $G_0$ across $F$. The corresponding map graph is unchanged, which contradicts the edge-maximality of $G_0$. Thus no face of $G_0$ is disconnected and therefore $G_0$ is connected.

Suppose that some face $F$ of $G_0$ has a repeated vertex $v$ in the boundary walk of $F$ and let $u,v,w$ be consecutive vertices in this boundary walk.  Since $v$ appears more than once in the boundary walk of $F$, the degree of $v$ is greater than $1$.  Since the degree of $v$ is greater than $1$ and the length of $F$ is greater than $2$, $u\neq w$. Since $v$ is repeated in the boundary walk of $F$, $uw$ is not an edge in the boundary of $f$. Add the edge $uw$ across $F$ and consider the resulting face $uvw$ to be a lake. Since $v$ appears elsewhere in the boundary of $F$, the corresponding map graph is unchanged, which contradicts the edge-maximality of $G_0$. Thus no facial walk of $G_0$ has a repeated vertex. Since each facial walk is connected, each face of $G_0$ is bounded by a cycle, and $G_0$ is biconnected.

@@

Let $G_1$ be a plane supergraph of $G_0^*$ obtained by adding the nation cycle of each face $F$ of $G_0$, and then triangulating each face with more than $d$ vertices on its boundary. $G_1$ is biconnected since $G^*_0$ is a biconnected spanning subgraph of $G_1$. Each nation cycle of length at least $3$ is now a facial cycle of $G_1$, and no face of $G_1$ has more than $d$ vertices on its boundary. Let $\widehat{G}_1$ be the edge-maximal $d$-framed graph whose frame is $G_1$.
